### 2025-05-20 preliminaries.tex
draft v1

This chapter describes an overview of the theoretical and formal theories required for the formalization of the Central Limit Theorem . This includes HOL Formalization, Measure Theory, Lebesgue Integration, and Probability Theory.
\section{HOL Formalization}
Higher Order Logic (HOL) \cite{hol4, slind2008brief} is derived from the Logic of Computable Functions (LCF) \cite{gordon1979edinburgh, milner1972logic} created by Robin Milner and colleagues in 1972. HOL is an adaptation of Church's Simple Theory of Types (STT) \cite{church1940formulation}, where a higher-order version of Hilbert's choice operator $\varepsilon $, Axiom of Infinity, and Rank-1 polymorphism have been added. HOL4 implements the original HOL framework, while other theorem provers in the HOL family, such as Isabelle/HOL, include important extensions. Such a simple logical basis makes HOL more accessible than those systems founded on much more advanced dependent type theories, such such the Calculus of Inductive and Co-Inductive Constructions constructed by Coq. Therefore, theories and proofs founded on HOL are easier for a layman to comprehend rather than being lost in a complicated type theory.

@@

\begin{enumerate}
    \item \textbf{Assumption Introduction (ASSUME):}
    \begin{itemize}
        \item Introduces a formula as an assumption.
        \item Rule:
        \item Example: From the assumption $P$, we conclude $P$.
    \end{itemize}

    \item \textbf{Reflexivity (REFL):}
    \begin{itemize}
        \item States that any term is equal to itself.
        \item Rule:
        \item Example: For $x : \texttt{int}$, $x = x$ is always true.
    \end{itemize}

    \item \textbf{Beta Conversion (BETA\_CONV):}
    \begin{itemize}
        \item Applies substitution in lambda abstractions.
        \item Rule:
        \item Example: $(\lambda x. x + 1)(5) \vdash 5 + 1$.
    \end{itemize}

    \item \textbf{Substitution (SUBST):}
    \begin{itemize}
        \item Replaces a term in a formula with another term proven to be equal.
        \item Rule:
        \item Example: From $x = y$ and $P(x)$, infer $P(y)$.
    \end{itemize}

    \item \textbf{Abstraction (ABS):}
    \begin{itemize}
        \item Generalizes an equation by abstracting a variable.
        \item Rule:
        \item Example: From $5 + 1 = 6$, infer $\lambda x. x + 1 = \lambda x. 6$.
    \end{itemize}

    \item \textbf{Type Instantiation (INST\_TYPE):}
    \begin{itemize}
        \item Specializes polymorphic functions or predicates to specific types.
        \item Rule:
        \item Example:
    \end{itemize}

    \item \textbf{Discharging Assumptions (DISCH):}
    \begin{itemize}
        \item Converts an assumption into an implication.
        \item Rule:
        \item Example: From $P \wedge Q$, infer $P \Rightarrow (Q \wedge P)$.
    \end{itemize}

    \item \textbf{Modus Ponens (MP):}
    \begin{itemize}
        \item Combines an implication and its premise to infer the conclusion.
        \item Rule:
        \item Example: From $x > 0 \Rightarrow x^2 > 0$ and $x > 0$, infer $x^2 > 0$.
    \end{itemize}
\end{enumerate}
These inference rules ensure that all logical derivations are traceable to basic axioms and established theorems. Additionally, the deductive system forms the backbone of HOL4, ensuring that proofs are both rigorous and reliable.

@@

\begin{enumerate}
    \item \textbf{Logical Truth ($T$)}
    \begin{itemize}
        \item Rule
        \item True is represented as the equality of two identical boolean functions.
    \end{itemize}

    \item \textbf{Logical Falsehood ($F$)}
    \begin{itemize}
        \item Rule
        \item False is defined to satisfy any boolean implication.
    \end{itemize}

    \item \textbf{Negation ($\neg$)}
    \begin{itemize}
        \item Rule
        \item Negation is the implication of a boolean value leading to falsehood.
    \end{itemize}

    \item \textbf{Conjunction ($\wedge$)}
    \begin{itemize}
        \item Rule
        \item Conjunction is defined as a logical function that evaluates nested implications.
    \end{itemize}

    \item \textbf{Conjunction ($\wedge$)}
    \begin{itemize}
        \item Rule
        \item Conjunction is defined as a logical function that evaluates nested implications.
    \end{itemize}

    \item \textbf{Disjunction ($\vee$)}
    \begin{itemize}
        \item Rule
        \item Disjunction is expressed through sequential implications.
    \end{itemize}

    \item \textbf{Universal Quantifier ($\forall$)}
    \begin{itemize}
        \item Rule
        \item Universality asserts that a predicate holds for all elements of a type.
    \end{itemize}

    \item \textbf{Existential Quantifier ($\exists$)}
    \begin{itemize}
        \item Rule
        \item Existence is defined using Hilbert’s choice operator ($\varepsilon$).
    \end{itemize}

@@

    \item \textbf{One-to-One ($One\_One$)}
    \begin{itemize}
        \item Rule
    \end{itemize}

    \item \textbf{Onto ($Onto$)}
    \begin{itemize}
        \item Rule
    \end{itemize}

HOL includes the constant $Type\_Definition$, which defines new types as bijections of subsets of existing types:
    \item \textbf{Type Definition $Type\_Definition$}
    \begin{itemize}
        \item Rule
        \item This process is automated by the HOL Datatype package, simplifying the creation of new types.

@@

HOL’s standard theory is built upon four foundational axioms:
\begin{enumerate}
    \item \textbf{Boolean Cases (BOOL\_CASES\_AX)}
    \begin{itemize}
        \item Rule
        \item This axiom ensures that any boolean value is either true or false.
    \end{itemize}

    \item \textbf{Eta Conversion (ETA\_AX)}
    \begin{itemize}
        \item Rule
        \item Eta conversion describes the extensionality of functions.
    \end{itemize}

    \item \textbf{Hilbert’s Choice (SELECT\_AX)}
    \begin{itemize}
        \item Rule
        \item This axiom relates the choice operator to existential quantification.
    \end{itemize}

    \item \textbf{Infinity (INFINITY\_AX)}
    \begin{itemize}
        \item Rule
        \item The Axiom of Infinity ensures the existence of an infinite set.
    \end{itemize}

@@

In HOL4, these components are represented as:
\begin{hol}
  \begin{alltt}
    Definition m\_space\_def :
    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)sp sts \(\mu\).
    m\_space (sp, sts, \(\mu\)) = sp
  \end{alltt}
\end{hol}

@@

\begin{hol}
  \begin{alltt}
    Definition measure\_def :
    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)sp sts \(\mu\).
    measure (sp, sts, \(\mu\)) = \(\mu\)
  \end{alltt}
\end{hol}

@@

A fundamental component of probability and integration in HOL4 is the notion of Borel sets and Borel-measurable functions. These are provided by two standard theories: \texttt{real\_borel}, which handles real-valued Borel sets and functions, and \texttt{borel}, which generalises these definitions to extended real values (\texttt{extreal}) and constructs Borel and Lebesgue measure spaces.

@@

Equivalently, \texttt{borel} can be generated using various families of half spaces or intervals, such as $ \{ x \mid x \leq a \} $, $ \{ x \mid a < x < b \} $, and so on.

@@

This construction allows measurable sets to include points at infinity, which is necessary when working with unbounded expectations or improper integrals.

\paragraph{Measurable Functions.}
A function $ f : X \to Y $ is measurable from a $\sigma$-algebra $ a $ to $ b $ if the preimage of any measurable set in $ b $ lies in the $\sigma$-algebra $ a $. This is formalised as:

@@

>>>>

This chapter describes an overview of the theoretical and formal theories required for the formalization of the Central Limit Theorem . This includes HOL Formalization, Measure Theory, Lebesgue Integration, and Probability Theory.
\section{HOL Formalization}
Higher Order Logic (HOL) \cite{hol4, slind2008brief} is derived from the Logic of Computable Functions (LCF) \cite{gordon1979edinburgh, milner1972logic} created by Robin Milner and colleagues in 1972. HOL is an adaptation of Church's Simple Theory of Types (STT) \cite{church1940formulation}, where a higher-order version of Hilbert's choice operator $\varepsilon $, Axiom of Infinity, and Rank-1 polymorphism have been added. HOL4 implements the original HOL framework, while other theorem provers in the HOL family, such as Isabelle/HOL, include important extensions. Such a simple logical basis makes HOL more accessible than those systems founded on much more advanced dependent type theories, such as the Calculus of Inductive and Co-Inductive Constructions constructed by Coq. Therefore, theories and proofs founded on HOL are easier for a layman to comprehend rather than being lost in a complicated type theory.

@@

\begin{enumerate}
    \item \textbf{Assumption Introduction (ASSUME):}
    \begin{itemize}
        \item Introduces a formula as an assumption.
        \item Rule:
          \begin{center}
          \begin{tabular}{c}
            \\ \hline
            $t${\small\verb+ |- +}$t$ \\
          \end{tabular}
        \end{center}
        \item Example: From the assumption $P$, we conclude $P$.
    \end{itemize}

    \item \textbf{Reflexivity (REFL):}
    \begin{itemize}
        \item States that any term is equal to itself.
        \item Rule:
          \begin{center}
            \begin{tabular}{c}
              \\ \hline
                 {\small\verb+ |- +}$t${\small\verb+ = +}$t$ \\
            \end{tabular}
          \end{center}
        \item Example: For $x : \texttt{int}$, $x = x$ is always true.
    \end{itemize}

    \item \textbf{Beta Conversion (BETA\_CONV):}
    \begin{itemize}
        \item Applies substitution in lambda abstractions.
        \item Rule:
          \begin{center}
            \begin{tabular}{c}
              \\ \hline
              \holtxt{ |- (\bs$x$.$t_1$)$t_2$ = $t_1[t_2/x]$}
            \end{tabular}
          \end{center}
        \item Example: $(\lambda x. x + 1)(5) \vdash 5 + 1$.
    \end{itemize}

    \item \textbf{Substitution (SUBST):}
    \begin{itemize}
        \item Replaces a term in a formula with another term proven to be equal.
        \item Rule:
          \begin{center}
            \begin{tabular}{c}
              $\Gamma_1${\small\verb+ |- +} $t_1${\small\verb+=+}$t'_1$ {\small\verb+  +} $\cdots$ {\small\verb+  +}
              $\Gamma_n${\small\verb+ |- +} $t_n${\small\verb+=+}$t'_n$ {\small\verb+  +}
              $\Gamma${\small\verb+ |- +} $t[t_1,\ldots,t_n]$ \\ \hline
              $\Gamma_1 \cup \cdots
              \cup \Gamma_n \cup \Gamma${\small\verb+ |- +} $t[t'_1,\ldots,t'_n]$ \\
            \end{tabular}
          \end{center}
        \item Example: From $x = y$ and $P(x)$, infer $P(y)$.
    \end{itemize}

    \item \textbf{Abstraction (ABS):}
    \begin{itemize}
        \item Generalizes an equation by abstracting a variable.
        \item Rule:
          \begin{center}
            \begin{tabular}{c}
              $\Gamma${\small\verb+ |- +}$t_1${\small\verb+ = +}$t_2$ \\ \hline
              $\Gamma${\small\verb+ |- (\+}$x${\small\verb+.+}$t_1${\small\verb+) = (\+}$x${\small\verb+.+}$t_2${\small\verb+)+} \\
            \end{tabular}
          \end{center}
        \item Example: From $5 + 1 = 6$, infer $\lambda x. x + 1 = \lambda x. 6$.
    \end{itemize}

    \item \textbf{Type Instantiation (INST\_TYPE):}
    \begin{itemize}
        \item Specializes polymorphic functions or predicates to specific types.
        \item Rule:
          \begin{center}
            \begin{tabular}{c}
              $\Gamma${\small\verb+ |- +}$t$ \\ \hline
              $\Gamma[\sigma_1,\ \ldots\ ,\sigma_n/\alpha_1,\ \ldots\ ,\alpha_n]${\small\verb+ |- +}$t[\sigma_1,\ \ldots\ ,\sigma_n/\alpha_1,\ \ldots\ ,\alpha_n]$
            \end{tabular}
          \end{center}
        \item Example: Consider the polymorphic identity theorem:
          $     \vdash \forall x.\ x = x
          $
          This holds for any type \texttt{$\alpha$}. Using \texttt{INST\_TYPE}, we can instantiate the type variable to a specific type, such as \texttt{real}. The result is:
          $     \vdash \forall x:\texttt{real}.\ x = x
          $
          This instantiated version now applies to real numbers specifically, which is often necessary in the context of formalising analysis or probability theory, where many theorems apply to real-valued functions or variables.
    \end{itemize}

    \item \textbf{Discharging Assumptions (DISCH):}
    \begin{itemize}
        \item Converts an assumption into an implication.
        \item Rule:
          \begin{center}
            \begin{tabular}{c}
              $\Gamma${\small\verb+ |- +} $t_2$ \\ \hline
              $\Gamma{-}\{t_1\}${\small\verb+ |- +} $t_1${\small\verb+ ==> +}$t_2$
            \end{tabular}
          \end{center}
        \item Example: From $P \wedge Q$, infer $P \Rightarrow (Q \wedge P)$.
    \end{itemize}

    \item \textbf{Modus Ponens (MP):}
    \begin{itemize}
        \item Combines an implication and its premise to infer the conclusion.
        \item Rule:
          \begin{center}
            \begin{tabular}{c}
              $\Gamma_1${\small\verb+ |- +}$t_1${\small\verb+ ==> +}$t_2$ {\small\verb+     +} $\Gamma_2${\small\verb+ |- +}$t_1$ \\
              \hline
              $\Gamma_1 \cup \Gamma_2${\small\verb+ |- +}$t_2$ \\
            \end{tabular}
          \end{center}
        \item Example: From $x > 0 \Rightarrow x^2 > 0$ and $x > 0$, infer $x^2 > 0$.
    \end{itemize}
\end{enumerate}
These inference rules ensure that all logical derivations are traceable to basic axioms and established theorems. Additionally, the deductive system forms the backbone of HOL4, ensuring that proofs are both rigorous and reliable.

@@

\begin{enumerate}
    \item \textbf{Logical Truth ($T$)}
    \begin{itemize}
    \item Rule
\begin{verbatim}
T_DEF              |- T  = ((\x:bool. x) = (\x. x))
\end{verbatim}
        \item True is represented as the equality of two identical boolean functions.
    \end{itemize}

    \item \textbf{Logical Falsehood ($F$)}
    \begin{itemize}
        \item Rule
\begin{verbatim}
F_DEF           |- F  = !t. t
\end{verbatim}
        \item False is defined to satisfy any boolean implication.
    \end{itemize}

    \item \textbf{Negation ($\neg$)}
    \begin{itemize}
        \item Rule
\begin{verbatim}
NOT_DEF         |- ~  = (\t. t ==> F)
\end{verbatim}
        \item Negation is the implication of a boolean value leading to falsehood.
    \end{itemize}

    \item \textbf{Conjunction ($\wedge$)}
    \begin{itemize}
        \item Rule
\begin{verbatim}
AND_DEF         |- /\ = \t1 t2. !t. (t1 ==> t2 ==> t) ==> t
\end{verbatim}
        \item Conjunction is defined as a logical function that evaluates nested implications.
    \end{itemize}

    \item \textbf{Disjunction ($\vee$)}
    \begin{itemize}
        \item Rule
\begin{verbatim}
OR_DEF          |- \/ = \t1 t2. !t. (t1 ==> t) ==> (t2 ==> t) ==> t
\end{verbatim}
        \item Disjunction is expressed through sequential implications.
    \end{itemize}

    \item \textbf{Universal Quantifier ($\forall$)}
    \begin{itemize}
        \item Rule
\begin{verbatim}
FORALL_DEF      |- !  = \P:'a->bool. P = (\x. T)
\end{verbatim}
        \item Universality asserts that a predicate holds for all elements of a type.
    \end{itemize}

    \item \textbf{Existential Quantifier ($\exists$)}
    \begin{itemize}
        \item Rule
\begin{verbatim}
EXISTS_DEF      |- ?  = \P:'a->bool. P($@ P)
\end{verbatim}
        \item Existence is defined using Hilbert’s choice operator ($\varepsilon$).
    \end{itemize}

@@

    \item \textbf{One-to-One ($One\_One$)}
    \begin{itemize}
        \item Rule
          \begin{alltt}
            \(\!\!\!{\turn}\!\!\!\!\) ONE_ONE = (\(\lambda\)f. \(\forall\)x1 x2. f x1 = f x2 \(\Rightarrow\) x1 = x2)
          \end{alltt}
    \end{itemize}

    \item \textbf{Onto ($Onto$)}
    \begin{itemize}
        \item Rule
          \begin{alltt}
            \(\!\!\!{\turn}\!\!\!\!\) ONTO = (\(\lambda\)f. \(\forall\)y. \(\exists\)x. y = f x)
          \end{alltt}
    \end{itemize}

HOL includes the constant $Type\_Definition$, which defines new types as bijections of subsets of existing types:
    \item \textbf{Type Definition (Type\_Definition)}
    \begin{itemize}
        \item Rule
          \begin{hol}
\begin{verbatim}
  |- TYPE_DEFINITION (P:'a->bool) (rep:'b->'a) =
  (!x' x''. (rep x' = rep x'') ==> (x' = x'')) /\
  (!x. P x = (?x'. x = rep x'))
\end{verbatim}
          \end{hol}
        \item This process is automated by the HOL Datatype package, simplifying the creation of new types.

@@

HOL’s standard theory is built upon four foundational axioms:
\begin{enumerate}
    \item \textbf{Boolean Cases (BOOL\_CASES\_AX)}
    \begin{itemize}
        \item Rule
          \begin{alltt}
            \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)t. (t \({\Leftrightarrow}\) T) \(\lor\) (t \({\Leftrightarrow}\) F)
          \end{alltt}
        \item This axiom ensures that any boolean value is either true or false.
    \end{itemize}

    \item \textbf{Eta Conversion (ETA\_AX)}
    \begin{itemize}
        \item Rule
          \begin{alltt}
            \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)t. (\(\lambda\)x. t x) = t
          \end{alltt}
        \item Eta conversion describes the extensionality of functions.
    \end{itemize}

    \item \textbf{Hilbert’s Choice (SELECT\_AX)}
    \begin{itemize}
        \item Rule
          \begin{alltt}
            \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)P x. P x \(\Rightarrow\) P (@ P)
          \end{alltt}
        \item This axiom relates the choice operator to existential quantification.
    \end{itemize}

    \item \textbf{Infinity (INFINITY\_AX)}
    \begin{itemize}
        \item Rule
          \begin{alltt}
            \(\!\!\!{\turn}\!\!\!\!\) \(\exists\)f. ONE_ONE f \(\land\) \(\neg\)ONTO f
          \end{alltt}
        \item The Axiom of Infinity ensures the existence of an infinite set.
    \end{itemize}

@@

In HOL4, these components are represented as:
\begin{center}
\scriptsize
\begin{hol}
  \begin{alltt}
    Definition m\_space\_def :
    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)sp sts \(\mu\).
    m\_space (sp, sts, \(\mu\)) = sp
  \end{alltt}
\end{hol}

@@

\begin{hol}
  \begin{alltt}
    Definition measure\_def :
    \(\!\!\!{\turn}\!\!\!\!\) \(\forall\)sp sts \(\mu\).
    measure (sp, sts, \(\mu\)) = \(\mu\)
  \end{alltt}
\end{hol}
\end{center}

@@

A further fundamental building block for probability and integration in HOL4 is the theory of Borel sets and Borel-measurable functions. These are provided by two standard theories: \texttt{real\_borel}, which considers real-valued Borel sets and functions, and \texttt{borel}, which extends these definitions to extended real values (\texttt{extreal}) and builds up Borel and Lebesgue measure spaces.

@@

Equivalently, \texttt{borel} can be generated using various families of half spaces or intervals, e.g $ \{ x \mid x \leq a \} $, $ \{ x \mid a < x < b \} $, and so on.

@@

The construction allows for measurable sets to have points at infinity, as required when working with unbounded expectations or improper integrals.

\paragraph{Measurable Functions.}
A function $f : X \to Y$ is measurable from a $\sigma$-algebra $a$ to $b$ if the preimage of every set in $b$ that is measurable is in the $\sigma$-algebra $a$.

Or, formally:

@@

HOL4 formalisation of Lebesgue integration provides a strong foundation for reasoning about expectations, variances, and convergence in probability. This section briefly introduces the key constructs and theorems that we use throughout the Central Limit Theorem formalisation.

\subsection{Positive Simple Functions and Integration}

Nonnegative measurable functions in HOL4 are approximated by positive simple functions, piecewise constant functions taking finitely many values on measurable disjoint sets. A function $f$ is represented as:

\[
f(x) = \sum_{j=1}^n y_j \cdot \mathbf{1}_{A_j}(x), \quad \text{where } A_j \text{ are disjoint measurable sets}.
\]

This is formalised in HOL4 using:
\begin{hol}
\begin{alltt}
Theorem pos\_simple\_fn\_def :
\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m f s a x.
pos\_simple\_fn m f s a x \({\Leftrightarrow}\)
(\(\forall\)t. t \(\in\) m\_space m \(\Rightarrow\) 0 \(\le\) f t) \(\land\)
(\(\forall\)t. t \(\in\) m\_space m \(\Rightarrow\)
        f t = \(\sum\) (\(\lambda\)i. Normal (x i) * \(\mathbbm{1}\) (a i) t) s) \(\land\)
(\(\forall\)i. i \(\in\) s \(\Rightarrow\) a i \(\in\) measurable\_sets m) \(\land\)
FINITE s \(\land\)
(\(\forall\)i. i \(\in\) s \(\Rightarrow\) 0 \(\le\) x i) \(\land\)
(\(\forall\)i j. i \(\in\) s \(\land\) j \(\in\) s \(\land\) i \(\ne\) j \(\Rightarrow\)
        DISJOINT (a i) (a j)) \(\land\)
BIGUNION (IMAGE a s) = m\_space m
\end{alltt}
\end{hol}

The integral of such a function is defined as:
\begin{hol}
\begin{alltt}
Definition pos\_simple\_fn\_integral\_def :
\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m s a x.
pos\_simple\_fn\_integral m s a x =
\(\sum\) (\(\lambda\)i. Normal (x i) * measure m (a i)) s
\end{alltt}
\end{hol}

\subsection{Positive and General Lebesgue Integral}

The positive integral of a nonnegative function $f$ is defined as the supremum of the integrals of all positive simple functions below $f$:
\begin{hol}
\begin{alltt}
Definition pos\_fn\_integral\_def :
\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m f.
\(\int\)\(\sp{+}\) m f =
sup \{r \,|\, (\(\exists\)g. r \(\in\) psfis m g \(\land\)
                \(\forall\)x. x \(\in\) m\_space m \(\Rightarrow\) g x \(\le\) f x)\}
\end{alltt}
\end{hol}

The general Lebesgue integral of a function $f$ is defined as:

\[
\int f = \int^+ f^+ - \int^+ f^-
\]
where $f^+ = \max(f, 0)$ and $f^- = \max(-f, 0)$.

\begin{hol}
\begin{alltt}
Definition integral\_def :
\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m f.
\(\int\) m f = \(\int\)\(\sp{+}\) m f\(\sp{+}\) \({-}\) \(\int\)\(\sp{+}\) m f\(\sp{-}\)
\end{alltt}
\end{hol}

\begin{hol}
\begin{alltt}
Definition fn\_plus\_def :
\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)f.
f\(\sp{+}\) = (\(\lambda\)x. if 0 < f x then f x else 0)
\end{alltt}
\end{hol}

\begin{hol}
\begin{alltt}
Definition fn\_minus\_def :
\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)f.
f\(\sp{-}\) = (\(\lambda\)x. if f x < 0 then -f x else 0)
\end{alltt}
\end{hol}

\subsection{Integrability}

A function is integrable if it is Borel-measurable and the integrals of both its positive and negative parts are finite:
\begin{hol}
\begin{alltt}
Definition integrable\_def :
\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m f.
integrable m f \({\Leftrightarrow}\)
f \(\in\) Borel\_measurable (measurable\_space m) \(\land\)
\(\int\)\(\sp{+}\) m f\(\sp{+}\) \(\ne\) +\(\infty\) \(\land\)
\(\int\)\(\sp{+}\) m f\(\sp{-}\) \(\ne\) +\(\infty\)
\end{alltt}
\end{hol}

\subsection{Almost Everywhere and Null Sets}

A property $P(x)$ holds almost everywhere (AE) if it holds outside of a null set:
\begin{hol}
\begin{alltt}
Definition AE\_DEF :
\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m P.
(AE x::m. P x) \({\Leftrightarrow}\)
\(\exists\)N. null\_set m N \(\land\)
\(\forall\)x. x \(\in\) m\_space m DIFF N \(\Rightarrow\) P x
\end{alltt}
\end{hol}

\subsection{Monotone Convergence Theorem}

In approximating general nonnegative functions by simple functions, the following theorem allows taking limits under the integral:
\begin{hol}
\begin{alltt}
Theorem lebesgue\_monotone\_convergence :
\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m f fi.
measure\_space m \(\land\)
(\(\forall\)i. fi i \(\in\) Borel\_measurable (measurable\_space m)) \(\land\)
(\(\forall\)i x. x \(\in\) m\_space m \(\Rightarrow\) 0 \(\le\) fi i x) \(\land\)
(\(\forall\)x. x \(\in\) m\_space m \(\Rightarrow\)
   mono\_increasing (\(\lambda\)i. fi i x)) \(\land\)
(\(\forall\)x. x \(\in\) m\_space m \(\Rightarrow\)
   sup (IMAGE (\(\lambda\)i. fi i x) \(\mathbb{U}\)(:num)) = f x) \(\Rightarrow\)
\(\int\)\(\sp{+}\) m f = sup (IMAGE (\(\lambda\)i. \(\int\)\(\sp{+}\) m (fi i)) \(\mathbb{U}\)(:num))
\end{alltt}
\end{hol}

\subsection{Fubini and Tonelli Theorems}

In formalising the Lindeberg replacement method, we often integrate over product spaces involving independent sequences of random variables. To justify iterated integrals in these settings, we rely on HOL4’s formalisation of Tonelli’s and Fubini’s Theorems, which ensure that integrals over product measure spaces can be computed as nested one-dimensional integrals.

\paragraph{Tonelli's Theorem.}
Tonelli's theorem applies to nonnegative functions and guarantees that the double integral equals the iterated integral regardless of whether the value is finite or infinite:

\begin{hol}
\begin{alltt}
Theorem TONELLI :
\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)X Y A B u v f.
sigma\_finite\_measure\_space (X,A,u) \(\land\)
sigma\_finite\_measure\_space (Y,B,v) \(\land\)
f \(\in\) Borel\_measurable ((X,A) \(\times\) (Y,B)) \(\land\)
(\(\forall\)s. s \(\in\) X \(\times\) Y \(\Rightarrow\) 0 \(\le\) f s) \(\Rightarrow\)
(\(\forall\)y. y \(\in\) Y \(\Rightarrow\) (\(\lambda\)x. f (x,y)) \(\in\) Borel\_measurable (X,A)) \(\land\)
(\(\forall\)x. x \(\in\) X \(\Rightarrow\) (\(\lambda\)y. f (x,y)) \(\in\) Borel\_measurable (Y,B)) \(\land\)
(\(\lambda\)x. \(\int\)\(\sp{+}\) (Y,B,v) (\(\lambda\)y. f (x,y))) \(\in\) Borel\_measurable (X,A) \(\land\)
(\(\lambda\)y. \(\int\)\(\sp{+}\) (X,A,u) (\(\lambda\)x. f (x,y))) \(\in\) Borel\_measurable (Y,B) \(\land\)
\(\int\)\(\sp{+}\) ((X,A,u) \(\times\) (Y,B,v)) f =
\(\int\)\(\sp{+}\) (Y,B,v) (\(\lambda\)y. \(\int\)\(\sp{+}\) (X,A,u) (\(\lambda\)x. f (x,y))) \(\land\)
\(\int\)\(\sp{+}\) ((X,A,u) \(\times\) (Y,B,v)) f =
\(\int\)\(\sp{+}\) (X,A,u) (\(\lambda\)x. \(\int\)\(\sp{+}\) (Y,B,v) (\(\lambda\)y. f (x,y)))
\end{alltt}
\end{hol}

\paragraph{Fubini's Theorem.}
Fubini’s theorem generalises Tonelli’s result to integrable (not necessarily nonnegative) functions. It justifies interchanging the order of integration under integrability assumptions:
\begin{hol}
\begin{alltt}
Theorem FUBINI :
\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)X Y A B u v f.
sigma\_finite\_measure\_space (X,A,u) \(\land\)
sigma\_finite\_measure\_space (Y,B,v) \(\land\)
f \(\in\) Borel\_measurable ((X,A) \(\times\) (Y,B)) \(\land\)
(\(\int\)\(\sp{+}\) ((X,A,u) \(\times\) (Y,B,v)) (abs \(\circ\) f) \(\ne\) +\(\infty\) \(\lor\)
 \(\int\)\(\sp{+}\) (Y,B,v) (\(\lambda\)y. \(\int\)\(\sp{+}\) (X,A,u) (\(\lambda\)x. (abs \(\circ\) f) (x,y))) \(\ne\) +\(\infty\) \(\lor\)
 \(\int\)\(\sp{+}\) (X,A,u) (\(\lambda\)x. \(\int\)\(\sp{+}\) (Y,B,v) (\(\lambda\)y. (abs \(\circ\) f) (x,y))) \(\ne\) +\(\infty\)) \(\Rightarrow\)
\(\int\)\(\sp{+}\) ((X,A,u) \(\times\) (Y,B,v)) (abs \(\circ\) f) \(\ne\) +\(\infty\) \(\land\)
\(\int\)\(\sp{+}\) (Y,B,v) (\(\lambda\)y. \(\int\)\(\sp{+}\) (X,A,u) (\(\lambda\)x. (abs \(\circ\) f) (x,y))) \(\ne\) +\(\infty\) \(\land\)
\(\int\)\(\sp{+}\) (X,A,u) (\(\lambda\)x. \(\int\)\(\sp{+}\) (Y,B,v) (\(\lambda\)y. (abs \(\circ\) f) (x,y))) \(\ne\) +\(\infty\) \(\land\)
integrable ((X,A,u) \(\times\) (Y,B,v)) f \(\land\)
(AE y::(Y,B,v). integrable (X,A,u) (\(\lambda\)x. f (x,y))) \(\land\)
(AE x::(X,A,u). integrable (Y,B,v) (\(\lambda\)y. f (x,y))) \(\land\)
integrable (X,A,u) (\(\lambda\)x. \(\int\) (Y,B,v) (\(\lambda\)y. f (x,y))) \(\land\)
integrable (Y,B,v) (\(\lambda\)y. \(\int\) (X,A,u) (\(\lambda\)x. f (x,y))) \(\land\)
\(\int\) ((X,A,u) \(\times\) (Y,B,v)) f =
\(\int\) (Y,B,v) (\(\lambda\)y. \(\int\) (X,A,u) (\(\lambda\)x. f (x,y))) \(\land\)
\(\int\) ((X,A,u) \(\times\) (Y,B,v)) f =
\(\int\) (X,A,u) (\(\lambda\)x. \(\int\) (Y,B,v) (\(\lambda\)y. f (x,y)))
\end{alltt}
\end{hol}

In this thesis, these theorems are applied in constructing auxiliary sequences of independent random variables from product spaces and evaluating their expectations. Their use guarantees that swapping integration order is mathematically and formally valid in our HOL4 development.

\subsection{$\mathcal{L}^p$ Spaces and Fundamental Inequalities}

In HOL4, the $\mathcal{L}^p$ space over a measure space $m$ is defined as the set of (extended) real-valued Borel-measurable functions whose $p$-th absolute power is integrable:
\begin{hol}
\begin{alltt}
Theorem lp\_space\_alt\_finite :
\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p m f.
0 < p \(\land\) p \(\ne\) +\(\infty\) \(\Rightarrow\)
(f \(\in\) lp\_space p m \({\Leftrightarrow}\)
 f \(\in\) Borel\_measurable (measurable\_space m) \(\land\)
 \(\int\)\(\sp{+}\) m (\(\lambda\)x. abs (f x) powr p) \(\ne\) +\(\infty\))
\end{alltt}
\end{hol}

The associated seminorm is defined as:

\[
\|f\|_p := \left( \int |f(x)|^p \, \mathrm{d}\mu(x) \right)^{1/p}
\]

\begin{hol}
\begin{alltt}
Theorem seminorm\_normal :
\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p m f.
0 < p \(\land\) p \(\ne\) +\(\infty\) \(\Rightarrow\)
seminorm p m f = \(\int\)\(\sp{+}\) m (\(\lambda\)x. abs (f x) powr p) powr p\(\sp{-}\)\(\sp{1}\)
\end{alltt}
\end{hol}

This seminorm forms the basis for all moment bounds and continuity results used in the CLT proof. In particular, the finiteness of third moments used in Lyapunov's condition corresponds to membership in $\mathcal{L}^3$.

\paragraph{Hölder’s Inequality.}
For functions $u \in \mathcal{L}^p(m)$ and $v \in \mathcal{L}^q(m)$ where $1/p + 1/q = 1$, Hölder’s inequality bounds their product:
\begin{hol}
\begin{alltt}
Theorem Hoelder\_inequality :
\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)m u v p q.
measure\_space m \(\land\) 0 < p \(\land\) 0 < q \(\land\)
p\(\sp{-}\)\(\sp{1}\) + q\(\sp{-}\)\(\sp{1}\) = 1 \(\land\)
u \(\in\) lp\_space p m \(\land\) v \(\in\) lp\_space q m \(\Rightarrow\)
integrable m (\(\lambda\)x. u x * v x) \(\land\)
\(\int\) m (\(\lambda\)x. abs (u x * v x)) \(\le\) seminorm p m u * seminorm q m v
\end{alltt}
\end{hol}

\paragraph{Minkowski’s Inequality.}
Also known as the triangle inequality for $\mathcal{L}^p$ norms:
\begin{hol}
\begin{alltt}
Theorem Minkowski\_inequality :
\(\!\!\!{\turn}\!\!\!\!\) \(\forall\)p m u v.
measure\_space m \(\land\) 1 \(\le\) p \(\land\)
u \(\in\) lp\_space p m \(\land\) v \(\in\) lp\_space p m \(\Rightarrow\)
(\(\lambda\)x. u x + v x) \(\in\) lp\_space p m \(\land\)
seminorm p m (\(\lambda\)x. u x + v x) \(\le\)
seminorm p m u + seminorm p m v
\end{alltt}
\end{hol}
